Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-05 talk/icooolps2011/paper.tex
add a sentence that the trace is SSA and stress where guards come from
implement Samuele's radical idea: just show the compounded speedups

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The line
numbers in the trace correspond to the line numbers in
Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
trace indeed contains
five calls to \texttt{dict.get}, which is slow. To make the language efficient
using a tracing JIT, we need to find a way to get rid of these dictionary
lookups somehow. How to achieve this will be topic of
Section~\ref{sec:fastobjmodel}.

@@

For space reasons we cannot perform a full evaluation here, but still want to
present some benchmark numbers. We chose to present two benchmarks, a port of
the classical Richards benchmark in Python and a Python version of the Telco
decimal benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
using a pure Python decimal floating point implementation. The results we see in
these two benchmarks seem to repeat themselves in other benchmarks using
object-oriented code; for purely numerical algorithms the speedups are 
significantly smaller.

@@

All benchmarks were run 50 times in the same process, to give the JIT time to
produce machine code. The arithmetic mean of the times of the last 30 runs were
used as the result. The errors were computed using a confidence interval with a
95\% confidence level \cite{georges_statistically_2007}. The results are
reported in Figure~\ref{fig:times}.

Versioned types speed up both benchmarks by a significant factor of around 7.
The speed advantage of maps alone is a lot less clear. Maps also have a memory
advantage which we did not measure here. By themselves, maps improved the
Richards benchmark slightly, but made the Telco benchmark slower. Enabling both
maps and versioned types together yields a significant improvement over just
versioned types for Richards. XXX good explanation. For Telco, enabling both
has little effect over the gains for versioned types alone.

\pedronis{XXX radical idea, given that there may be no space to discuss the subtle points and that this is really about showing that we can enable such mechanisms and there is already literature that shows that/how they work, to just consider the benchmarks with no maps and no versions and with both?}

\begin{figure}
{\footnotesize
\begin{center}
\begin{tabular}{|l|r|r|}
\hline
        &richards[ms]       &telco[ms] \\
\hline
CPython &357.79 $\pm$ 1.32     &1209.67 $\pm$ 2.20\\
speedup &1.00 $\times$ &1.00 $\times$\\
\hline
JIT baseline    &421.87 $\pm$ 0.48     &738.18 $\pm$ 3.29\\
speedup &0.85 $\times$ &1.64 $\times$\\
\hline
JIT map &382.88 $\pm$ 4.40     &834.19 $\pm$ 4.91\\
speedup &0.93 $\times$ &1.45 $\times$\\
\hline
JIT version     &49.87 $\pm$ 0.29      &157.88 $\pm$ 1.79    \\
speedup &7.17 $\times$ &7.66 $\times$\\
\hline
JIT full        &17.89 $\pm$ 1.15      &153.48 $\pm$ 1.86    \\
speedup &20.00 $\times$        &7.88 $\times$\\
\hline
\end{tabular}
\end{center}
}
\caption{Benchmark Results}
\label{fig:times}
\end{figure}

>>>>

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The line
numbers in the trace correspond to the line numbers in
Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
trace is in SSA form. Note how all the guards in trace correspond to a
condition in the original code. The trace contains
five calls to \texttt{dict.get}, which is slow. To make the language efficient
using a tracing JIT, we need to find a way to get rid of these dictionary
lookups somehow. How to achieve this will be topic of
Section~\ref{sec:fastobjmodel}.

@@

For space reasons we cannot perform a full evaluation here, but still want to
present some benchmark numbers. We chose to present just some benchmarks: The
templating engine of the Django web
framework\footnote{\texttt{http://www.djangoproject.com/}}; a Monte-Carlo Go
AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/
disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
Richards benchmark in Python; a Python version of the Telco decimal
benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
pure Python decimal floating point implementation. The results we see in these
benchmarks seem to repeat themselves in other benchmarks using object-oriented
code; for purely numerical algorithms the speedups are significantly smaller.

@@

All benchmarks were run 50 times in the same process, to give the JIT time to
produce machine code. The arithmetic mean of the times of the last 30 runs were
used as the result. The errors were computed using a confidence interval with a
95\% confidence level \cite{georges_statistically_2007}. The results are
reported in Figure~\ref{fig:times}, together with the same numbers normed to
those of the full JIT.

The optimizations give a speedup between 80\% and almost 20 times. The Richards
is a particularly good case for the optimizations as it makes heavy uses of
object-oriented features. Pyflate uses mostly imperative code, so does not
benefit as much. Together with the optimization, PyPy outperforms CPython in
all benchmarks, which is not surprising because CPython is a simple
bytecode-based interpreter.

\begin{figure}
\begin{center}
{\footnotesize
\begin{tabular}{|l|r|r|r|}
\hline
	&CPython	&JIT baseline	&JIT full\\
\hline
django[ms]	&988.67 $\pm$ 0.49	&405.62 $\pm$ 4.80	&149.31 $\pm$ 1.37\\
	&6.62 $\times$	&2.72 $\times$	&1.00 $\times$\\
\hline
go[ms]	&947.43 $\pm$ 1.30	&525.53 $\pm$ 7.67	&174.32 $\pm$ 7.78\\
	&5.44 $\times$	&3.01 $\times$	&1.00 $\times$\\
\hline
pyflate[ms]	&3209.20 $\pm$ 3.65	&2884.26 $\pm$ 21.11	&1585.48 $\pm$ 5.22\\
	&2.02 $\times$	&1.82 $\times$	&1.00 $\times$\\
\hline
richards[ms]	&357.79 $\pm$ 1.32	&421.87 $\pm$ 0.48	&17.89 $\pm$ 1.15\\
	&20.00 $\times$	&23.58 $\times$	&1.00 $\times$\\
\hline
telco[ms]	&1209.67 $\pm$ 2.20	&738.18 $\pm$ 3.29	&153.48 $\pm$ 1.86\\
	&7.88 $\times$	&4.81 $\times$	&1.00 $\times$\\
\hline
\end{tabular}
}
\end{center}
\caption{Benchmark Results}
\label{fig:times}
\end{figure}
